Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-09 application.tex
paper updates, code updates

\subsection{Newtonian cooling}\label{newton_cool}
For $l\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{lin_energy}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$L\ll l_\mathrm{rad}$.

@@

The most unstable mode occurs at $\khat\simeq30$ with growth rate 
\begin{align*}
  \nu =0.058\epsilon\Omega_k = 3.9\times10^{-3}\Omega_k.  
\end{align*}
This corresponds to a characteristic growth time $\sim
40$ orbits. However, growth rates are weakly depedent on $\khat$ for
$10\lesssim \khat 60$ so this may be taken as a representative growth
rate of the fundamental VSI in the MMSN at $r=50\mathrm{AU}$. The
eigenfunction for $\khat=30$ is shown in Fig. \ref{mmsn_eigenW}.

>>>>

\subsection{Newtonian cooling}\label{newton_cool}
For $l\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right), 
\end{align}
where $c_v$ is the heat capacity at constant volume. When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{full_energy1}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$l\ll l_\mathrm{rad}$.

@@

The most unstable mode occurs at $\khat\simeq30$ with growth rate 
\begin{align*}
  \nu =0.058\epsilon\Omega_k = 3.9\times10^{-3}\Omega_k.  
\end{align*}
This corresponds to a characteristic growth time $\sim
40$ orbits. However, since growth rates are weakly depedent on $\khat$ for
$10\lesssim \khat \lesssim 60$, this may be taken as a representative growth
time of the fundamental VSI in the MMSN at $r=50\mathrm{AU}$. The
eigenfunction for $\khat=30$ is shown in Fig. \ref{mmsn_eigenW}.

### 2015-02-09 linear_iso.tex
paper updates, code updates

The linear problem in the VGSB framework can be derived from 
Eq. \ref{lin_mass}---\ref{lin_vz} by setting $\hat{g}_c=0$, since 
the VGSB, like the standard shearing box, is homogeneous in the radial 
direction in equilibrium. The resulting linear problem is very similar
to that for local axisymmetric waves in accretion disks already
discussed by \cite{lubow93}, but modified by the presence of vertical
shear and thermal relaxation. This is convenient as it allows us to
directly apply their solution methods.

>>>>

The linear problem in the VGSB framework can be derived from 
Eq. \ref{lin_mass}---\ref{lin_energy} by setting $\hat{g}_c=0$, since 
the VGSB, like the standard shearing box, is homogeneous in the radial 
direction in equilibrium. The resulting linear problem is very similar
to that for local axisymmetric waves in accretion disks already
discussed by \cite{lubow93}, but modified by the presence of vertical
shear and thermal relaxation. This is convenient as it allows us to
directly apply their solution methods.

### 2015-02-09 setup.tex
paper updates, code updates

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. The equilibrium disk satisfies
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} 
%where $\Phi_* = -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational
%potential of the central star.
We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H_\mathrm{iso}}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant, and $H_\mathrm{iso}$ is a
characteristic scale-height.  Thus, $\epsilon$ is also the
characteristic disk aspect-ratio.

@@

The linearized fluid equations in cylindrical co-ordinates are  
\begin{align}
  \ii\sigma\delta\rho &= \frac{\rho}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p}{\p
    z}\left(\rho\delta v_z\right)  + \hat{g}_c\frac{\p\rho}{\p r}\delta v_r,\\
  \ii\sigma\delta v_r &= \frac{1}{\rho}\frac{\p\delta P}{\p r} - 2\Omega
  \delta v_\phi - \hat{g}_c\frac{\delta\rho}{\rho^2}\frac{\p P}{\p r},\\ 
  \ii\sigma\delta v_\phi  &= \frac{\kappa^2}{2\Omega}\delta v_r +
  r\frac{\p\Omega}{\p z}\delta v_z,\\
  \ii \sigma\delta v_z &= \frac{1}{\rho}\frac{\p\delta P}{\p z} -
  \frac{\delta \rho}{\rho^2}\frac{\p P}{\p z},\\
  \ii\sigma\delta P &=\frac{\p P}{\p z}\delta v_z + \gamma P
  \left[\frac{1}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p
      \delta v_z}{\p z}\right]\notag \\
  &+ \frac{1}{t_c}\left(\delta P -  
    \frac{P}{\rho}\delta\rho\right) + \hat{g}_c \frac{\p P}{\p r}\delta v_r,
\end{align}
where the coefficients $\hat{g}_c=1$ are inserted to label terms
associated with the background radial disk structure.

>>>>

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. The equilibrium disk satisfies
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} 
%where $\Phi_* = -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational
%potential of the central star.
We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H_\mathrm{iso}}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k$ is the Keplerian azimuthal velocity and
$H_\mathrm{iso}$ is a characteristic scale-height.  Thus, $\epsilon$
is also the characteristic disk aspect-ratio.

@@

The linearized fluid equations in cylindrical co-ordinates are  
\begin{align}
  \ii\sigma\delta\rho &= \frac{\rho}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p}{\p
    z}\left(\rho\delta v_z\right)  + \hat{g}_c\frac{\p\rho}{\p r}\delta v_r,\\
  \ii\sigma\delta v_r &= \frac{1}{\rho}\frac{\p\delta P}{\p r} - 2\Omega
  \delta v_\phi - \hat{g}_c\frac{\delta\rho}{\rho^2}\frac{\p P}{\p r},\\ 
  \ii\sigma\delta v_\phi  &= \frac{\kappa^2}{2\Omega}\delta v_r +
  r\frac{\p\Omega}{\p z}\delta v_z,\\
  \ii \sigma\delta v_z &= \frac{1}{\rho}\frac{\p\delta P}{\p z} -
  \frac{\delta \rho}{\rho^2}\frac{\p P}{\p z},\\
  \ii\sigma\delta P &=\frac{\p P}{\p z}\delta v_z + \gamma P
  \left[\frac{1}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p
      \delta v_z}{\p z}\right]\notag \\
  &+ \frac{1}{t_c}\left(\delta P -  
    \frac{P}{\rho}\delta\rho\right) + \hat{g}_c \frac{\p P}{\p r}\delta v_r,\label{full_energy1}
\end{align}
where the coefficients $\hat{g}_c=1$ are inserted to label terms
associated with the background radial disk structure.
